# (PART) ALPHA DIVERSITY {-}

# Alpha Diversity Analysis
> The input files must contain the raw read count per sample (only integers).
- Taxa must be in rows and if not transpose the table.

## Import Count Data
```{r }
source("R/common.R")
load("RDataRDS/taxa_abund_tables.RData", verbose = T)

```
## Kingdom Richness
```{r}
k_richness <- df_kingdom %>% tibble::column_to_rownames("Kingdom") %>% mutate_if(is.numeric, ~1 * (. > 0))
k_richness[, 1:5]

```

## Phylum Richness
```{r}
p_richness <- df_phylum %>% tibble::column_to_rownames("Phylum") %>% mutate_if(is.numeric, ~1 * (. > 0))
p_richness[1:5, 1:5]

```

## Class Richness
```{r}
c_richness <- df_class %>% tibble::column_to_rownames("Class") %>% mutate_if(is.numeric, ~1 * (. > 0))
c_richness[1:5, 1:5]

```

## Order Richness
```{r}
o_richness <- df_order %>% tibble::column_to_rownames("Order") %>% mutate_if(is.numeric, ~1 * (. > 0))
o_richness[1:5, 1:5]

```

## Family Richness
```{r}
f_richness <- df_family %>% tibble::column_to_rownames("Family") %>% mutate_if(is.numeric, ~1 * (. > 0))
f_richness[1:5, 1:5]

```

## Genus Richness
```{r}
g_richness <- df_genus %>% tibble::column_to_rownames("Genus") %>% mutate_if(is.numeric, ~1 * (. > 0))
g_richness[1:5, 1:5]

```

## Species Richness
```{r}
s_richness <- df_species %>% tibble::column_to_rownames("Species") %>% mutate_if(is.numeric, ~1 * (. > 0))
s_richness[1:5, 1:5]

```

## Cummulative Taxa Richness

```{r}
total_richness <- rbind(
  Kingdom = k_richness %>% mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>% colSums(),
  Phylum = p_richness %>% mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>% colSums(),
  Class = c_richness %>% mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>% colSums(),
  Order = o_richness %>% mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>% colSums(),
  Family = f_richness %>% mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>% colSums(),
  Genus = g_richness %>% mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>% colSums(),
  Species = s_richness %>% mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>% colSums())
  
  total_richness <- total_richness %>% dfRowName("clade_name")
  total_richness

write_csv(total_richness, "RDataRDS/total_richness.csv" )
saveRDS(total_richness, "RDataRDS/total_richness.rds" )
```


## Richness Bar chart
```{r richness_bar, fig.height=10, fig.width=7, include=FALSE}
richness_plot <- total_richness %>% reshape2::melt() %>% 
  mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>% 
  filter(value > 0.0) %>%
  ggplot(aes(x=reorder(variable, -value), y = value, fill = value), color=clade_name) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    x = "\nSample", 
    y = "Species Richness", 
    fill = "Richness") +
  theme(
    axis.text = element_text(hjust = 0.1, size = 12),
    axis.title=element_text(size=14,face="bold"),
    axis.text.x = element_text(hjust=0.5, angle = 90),
    legend.position = "right") +
    guides(fill=guide_legend(ncol=1)) +
  theme_bw() + 
  coord_flip()
richness_plot
```

```{r richness_bar_gif, echo=FALSE, fig.height=5, fig.width=7, animation.hook='gifski', interval=3, fig.path="./GIFS/"}
richness_plot
```

```{r}
# Richness R Objects
save(
  # taxon-level and total richness
  k_richness, 
  p_richness, 
  c_richness, 
  o_richness, 
  f_richness, 
  g_richness, 
  s_richness, 
  total_richness, 
  richness_plot,

  file = "RDataRDS/taxa_richness.RData" )

```

## Diversity index
```{r fig.height=7, fig.width=5}
# source("R/alpha.R")
library(vegan)
library(tidyverse)

set.seed(2022)

taxlevels <- readRDS("RDataRDS/taxlevels.rds")
days_wanted <- c(0:9, 141:150)

shared <- readRDS("RDataRDS/otu_taxa_abund.rds") %>%
  select(-taxlevels) %>%
  distinct(OTU, .keep_all = T) %>% 
  mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>% 
  data.table::transpose(keep.names = "Group", make.names = "OTU") %>% 
  pivot_longer(-Group) %>%
  group_by(Group) %>%
  mutate(total = sum(value)) %>%
  filter(total > 1800) %>%
  group_by(name) %>%
  mutate(total = sum(value)) %>%
  filter(total != 0) %>%
  ungroup() %>%
  select(-total)

rand <- shared %>%
  uncount(value) %>%
  mutate(name = sample(name)) %>%
  count(Group, name, name="value")

richness <- function(x){
  
  # r <- sum(x > 0)
  # return(r)
  
  sum(x>0)
}

shannon <- function(x){
  
  rabund <- x[x>0]/sum(x)
  -sum(rabund * log(rabund))
  
}

simpson <- function(x){
  
  n <- sum(x)
  
  # sum(x * (x-1) / (n * (n-1)))
  1 - sum((x/n)^2)
}

library(vegan)
rand %>%
  group_by(Group) %>%
  summarize(sobs = specnumber(value),
            shannon = diversity(value, index="shannon"),
            simpson = diversity(value, index="simpson"),
            invsimpson = 1/simpson,
            n = sum(value)) %>%
  pivot_longer(cols=c(sobs, shannon, invsimpson, simpson),
               names_to="metric") %>%
  ggplot(aes(x=n, y=value)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~metric, nrow=4, scales="free_y") + xbreaks5

```


```{r}
## Current saved objects
resave(
  # taxon-level and total richness
  k_richness, 
  p_richness, 
  c_richness, 
  o_richness, 
  f_richness, 
  g_richness, 
  s_richness, 
  total_richness,
  richness_plot,
  file = "RDataRDS/project6data.RData")

load("RDataRDS/project6data.RData", verbose = T)

```
