# (PART) TASK 2: 6-11 FILTER REL ABUND {-}

# Filter Abundance Less than 0.005  {#task-6-11}

## Import and processed data
```{r}
data <- readRDS("RDataRDS/taxlevels.rds")
otu_taxa_abund <- readRDS("RDataRDS/otu_taxa_abund.rds")

data <- otu_taxa_abund[, -1] %>% select(Species, !all_of(taxlevels)) %>% 
  distinct(Species, .keep_all = T) %>% tibble::column_to_rownames("Species")
head(data)[1:5, 1:5]

```

## Prepare dataset1 to filter from
- Dataset 1 is unfiltered
- Abundance is set at 0.0%
- Prevalence is set at 0%
```{r}
library(OTUtable)

dataset1 <- filter_taxa(data, abundance = 0.0, persistence = 0)

cat("Dimension of dataset1\n")
dim(dataset1)

cat("Dataset1 structure\n\n")
dataset1[1:5, 1:5]

saveRDS(dataset1, "RDataRDS/dataset1.rds")
```

## Prepare dataset 2
- Data set 2 contains species with a relative abundance of 0.005 or more
- Abundance is set at 0.005%
- Prevalence is set at 0%
```{r}
library(OTUtable)

dataset2 <- filter_taxa(dataset1, abundance = 0.005, persistence = 0)

cat("Dimension of dataset2\n")
dim(dataset2)

cat("Dataset2 structure\n\n")
dataset2[1:5, 1:5]

saveRDS(dataset2, "RDataRDS/dataset2.rds")
```


# (PART) TASK 2: 12-17 FILTER REL ABUND {-}

# Filter abundance less than 0.005 and prevalence less than 10 perc {#task-12-17}
- Data set 3 contains species with a relative abundance of 0.005 or more and occurring in 10% (or more) of all samples.

## Prepare dataset 3
```{r}
dataset2 <- readRDS("RDataRDS/dataset2.rds")
dataset3 <- filter_taxa(dataset2, abundance = 0.005, persistence = 10)
cat("Dimension of dataset3\n")
dim(dataset3)

cat("Dataset3 structure\n\n")
dataset3[1:5, 1:5]

saveRDS(dataset2, "RDataRDS/dataset3.rds")
```
