# Data Distribution {#data-visuals}

```{r}
## Load libraries and data
source("R/common.R")
library(tidyverse)
library(ggtext)
library(phyloseq)
library(metagMisc)
library(microbiome)
library(microViz)
library(RColorBrewer)
library(glue)
library(broom)

# View object content
load("../iMAP-part3/RDataRDS/Rjoined_objects.RData", verbose = T)
load("../iMAP-part3/RDataRDS/phylo_taxonomy_objects.RData", verbose = T)
```

> Note: Some analyses may introduce infinite values in continuous x-axis for zero relative abundance. The rows containing non-finite values will be removed. Solution is replace the zeros with a tiny non-significant value below the LoD (limit od detection).

## Computing limit of detection (LoD)
```{r}
set.seed(110912)

nseqs_per_sample <- Rjoined_sub_object %>% 
  group_by(sample_id) %>%
  summarise(N = sum(count), .groups="drop") %>%
  count(N) %>%
  pull(N)

stopifnot(length(nseqs_per_sample) == 1)

lod <- max(100* 1/nseqs_per_sample)
lod
```


## Prepare taxa abundance data {#taxa-abundance}

### Relative abundance
```{r}
## kingdom
otu_rel_abund <- Rjoined_sub_object %>% 
  filter(level == "kingdom",
         group == "Buffalo" | group == "Wildebeest",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)
         ) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop")

taxon_pool <- otu_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(rel_abund) < 0,
            median = median(rel_abund),
            .groups = "drop")

kingdom_rel_abund <- inner_join(otu_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = 100 * sum(rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0))

#-------phylum----------------
otu_rel_abund <- Rjoined_sub_object %>% 
  filter(level == "phylum",
         group == "Buffalo" | group == "Wildebeest",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)
         ) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop")

taxon_pool <- otu_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(rel_abund) < 2,
            median = median(rel_abund),
            .groups = "drop")

phylum_rel_abund <- inner_join(otu_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = 100 * sum(rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0))

#-------class----------------
otu_rel_abund <- Rjoined_sub_object %>% 
  filter(level == "class",
         group == "Buffalo" | group == "Wildebeest",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)
         ) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop")

taxon_pool <- otu_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(rel_abund) < 2,
            median = median(rel_abund),
            .groups = "drop")

class_rel_abund <- inner_join(otu_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = 100 * sum(rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0))

#-------order----------------
otu_rel_abund <- Rjoined_sub_object %>% 
  filter(level == "order",
         group == "Buffalo" | group == "Wildebeest",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)
         ) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop")

taxon_pool <- otu_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(rel_abund) < 2,
            median = median(rel_abund),
            .groups = "drop")

order_rel_abund <- inner_join(otu_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = 100 * sum(rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0))

#-------family----------------
otu_rel_abund <- Rjoined_sub_object %>% 
  filter(level == "family",
         group == "Buffalo" | group == "Wildebeest",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)
         ) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop")

taxon_pool <- otu_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(rel_abund) < 2,
            median = median(rel_abund),
            .groups = "drop")

family_rel_abund <- inner_join(otu_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = 100 * sum(rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0))

#-------genus----------------
otu_rel_abund <- Rjoined_sub_object %>% 
  filter(level == "genus",
         group == "Buffalo" | group == "Wildebeest",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)
         ) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop")

taxon_pool <- otu_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(rel_abund) < 2,
            median = median(rel_abund),
            .groups = "drop")

genus_rel_abund <- inner_join(otu_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = 100 * sum(rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0))

save(kingdom_rel_abund, 
     phylum_rel_abund, 
     class_rel_abund, 
     order_rel_abund, 
     family_rel_abund, 
     genus_rel_abund,
     file = "RDataRDS/taxa_abund_data.RData")
```


### Median relative abundance
```{r sample_barchat_fig, fig.height=4, fig.width=6, animation.hook='gifski', interval=4}
## kingdom
otu_rel_abund <- Rjoined_sub_object %>% 
  filter(level == "kingdom",
         group == "Buffalo" | group == "Wildebeest",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)
         ) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- otu_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 2,
            median = median(median_rel_abund),
            .groups = "drop")

kingdom_median_rel_abund <- inner_join(otu_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0))

#-------phylum----------------
otu_rel_abund <- Rjoined_sub_object %>% 
  filter(level == "phylum",
         group == "Buffalo" | group == "Wildebeest",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)
         ) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- otu_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 2,
            median = median(median_rel_abund),
            .groups = "drop")

phylum_median_rel_abund <- inner_join(otu_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0))

#-------class----------------
otu_rel_abund <- Rjoined_sub_object %>% 
  filter(level == "class",
         group == "Buffalo" | group == "Wildebeest",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)
         ) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- otu_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 3,
            median = median(median_rel_abund),
            .groups = "drop")

class_median_rel_abund <- inner_join(otu_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0))

#-------order----------------
otu_rel_abund <- Rjoined_sub_object %>% 
  filter(level == "order",
         group == "Buffalo" | group == "Wildebeest",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)
         ) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- otu_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 2,
            median = median(median_rel_abund),
            .groups = "drop")

order_median_rel_abund <- inner_join(otu_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0))

#-------family----------------
otu_rel_abund <- Rjoined_sub_object %>% 
  filter(level == "family",
         group == "Buffalo" | group == "Wildebeest",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)
         ) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- otu_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 3,
            median = median(median_rel_abund),
            .groups = "drop")

family_median_rel_abund <- inner_join(otu_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0))

#-------genus----------------
otu_rel_abund <- Rjoined_sub_object %>% 
  filter(level == "genus",
         group == "Buffalo" | group == "Wildebeest",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*", taxon)
         ) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- otu_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 2,
            median = median(median_rel_abund),
            .groups = "drop")

genus_median_rel_abund <- inner_join(otu_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0))

#-------genus_glued----------------
otu_rel_abund <- Rjoined_sub_object %>% 
  select(-taxon) %>% 
  rename(taxon = taxon_phylo) %>% 
  filter(level == "genus",
         group == "Buffalo" | group == "Wildebeest",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*", taxon)
         ) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- otu_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 2,
            median = median(median_rel_abund),
            .groups = "drop")

genus_glued_median_rel_abund <- inner_join(otu_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0))

resave(kingdom_median_rel_abund, 
     phylum_median_rel_abund,  
     class_median_rel_abund, 
     order_median_rel_abund,  
     family_median_rel_abund,  
     genus_median_rel_abund,
     genus_glued_median_rel_abund,
     file = "RDataRDS/taxa_abund_data.RData")
```


## Stacked barchart across samples

```{r}
load("RDataRDS/taxa_abund_data.RData", verbose = T)
```

```{r stacked_bar_fig, fig.height=4, fig.width=6, animation.hook='gifski', interval=4}
kingdom_median_rel_abund %>%
  ggplot(aes(x = sample_id, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Relative Abundance (%)",
      subtitle = "Stacked barchart filled by kingdom",
       fill = NULL) +
  theme_classic() +
  theme(legend.text = element_markdown(),
        legend.key.size = unit(10, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip()


phylum_median_rel_abund %>%
  ggplot(aes(x = sample_id, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Relative Abundance (%)",
      subtitle = "Stacked barchart filled by phylum",
       fill = NULL) +
  theme_classic() +
  theme(legend.text = element_markdown(),
        legend.key.size = unit(10, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip()


class_median_rel_abund %>%
  ggplot(aes(x = sample_id, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Relative Abundance (%)",
      subtitle = "Stacked barchart filled by class",
       fill = NULL) +
  theme_classic() +
  theme(legend.text = element_markdown(),
        legend.key.size = unit(10, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip()


order_median_rel_abund %>%
  ggplot(aes(x = sample_id, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Relative Abundance (%)",
      subtitle = "Stacked barchart filled by order",
       fill = NULL) +
  theme_classic() +
  theme(legend.text = element_markdown(),
        legend.key.size = unit(10, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip()


family_median_rel_abund %>%
  ggplot(aes(x = sample_id, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Relative Abundance (%)",
      subtitle = "Stacked barchart filled by family",
       fill = NULL) +
  theme_classic() +
  theme(legend.text = element_markdown(),
        legend.key.size = unit(10, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip()

genus_median_rel_abund %>%
  ggplot(aes(x = sample_id, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Relative Abundance (%)",
      subtitle = "Stacked barchart filled by genus",
       fill = NULL) +
  theme_classic() +
  theme(legend.text = element_markdown(),
        legend.key.size = unit(10, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip()
```


## Exploring genera relative abundance {#genus_abundance}

## Grouped barchart
```{r}
genus_median_rel_abund  <- genus_glued_median_rel_abund
genus_median_rel_abund %>%
  ggplot(aes(x = taxon, y = median_rel_abund, fill = group)) +
  geom_col(width=0.8, position = position_dodge()) +
  labs(x = NULL,
       y = "Relative Abundance (%)",
      subtitle = "Grouped barchart filled by group",
       fill = NULL) +
  theme_classic() +
  theme(axis.text.x = element_markdown(angle = 0, hjust = 1, vjust = 1),
        axis.text.y = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        panel.background = element_blank(),
        panel.grid.major.x =  element_line(colour = "lightgray", size = 0.1),
        panel.border = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip()
```


## Jitter strip chart
```{r}
genus_median_rel_abund %>%
  ggplot(aes(y = taxon, x = median_rel_abund, fill = group)) +
  geom_jitter(position = position_jitterdodge(
    jitter.width = 0.3, 
    dodge.width = 0.4),
    pch = 21, 
    stroke = 0,
    size = 1.8) +
  labs(
    y = NULL,
    x = "Relative Abundance (%)",
    title = "Strip Plot",
    fill = NULL) +
    theme_classic() +
    theme(
      axis.text.y = element_markdown(angle = 0, hjust = 1, vjust = 1),
      legend.key.size = unit(12, "pt"),
      panel.background = element_blank(),
      panel.grid.major.y =  element_line(colour = "lightgray", size = 0.1),
      panel.border = element_blank()) +
    guides(fill = guide_legend(ncol=1)) +
    scale_x_continuous(expand = c(0, 0))
```


## Boxplot

```{r}
genus_median_rel_abund %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(y = taxon, x = median_rel_abund, color = group)) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.8)) +
  # geom_jitter(position = position_jitterdodge(
  #   jitter.width = 0.3,
  #   dodge.width = 0.4),
  #   pch = 19) +
    # scale_color_manual(name = NULL,
    #                 breaks = c("Buffalo", "Wildebeest"),
    #                 labels = c("Buffalo", "Wildebeest"),
    #                 values = c("blue4", "red")) +
    labs(
    y = NULL,
    x = "Relative Abundance (%)",
    title = "Simple Boxplot",
    color = NULL) +
    theme_classic() +
    theme_classic() +
  theme(axis.text.y = element_markdown(angle = 0, hjust = 1, vjust = 1),
        # axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        legend.background = element_rect(colour = "gray", fill = NA),
        legend.margin = margin(t = 5, r = 5, b = 5, l = 5)) +
  guides(fill = guide_legend(ncol=1)) +
  scale_x_continuous()
```


## Jitter and Boxplot
```{r}
genus_median_rel_abund %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(y = taxon, x = median_rel_abund, color = group)) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.8)) +
  geom_jitter(position = position_jitterdodge(
    jitter.width = 0.3,
    dodge.width = 0.4),
    pch = 19) +
    # scale_color_manual(name = NULL,
    #                 breaks = c("Buffalo", "Wildebeest"),
    #                 labels = c("Buffalo", "Wildebeest"),
    #                 values = c("blue4", "red")) +
    labs(
    y = NULL,
    x = "Relative Abundance (%)",
    title = "Box and jitter plot",
    color = NULL) +
    theme_classic() +
    theme_classic() +
  theme(axis.text.y = element_markdown(angle = 0, hjust = 1, vjust = 1),
        # axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        legend.background = element_rect(colour = "gray", fill = NA),
        legend.margin = margin(t = 5, r = 5, b = 5, l = 5)) +
  guides(fill = guide_legend(ncol=1)) +
  scale_x_continuous()
```



## Point range with log scaled axis {#log-scaled}
```{r}
otuplot <- genus_median_rel_abund %>% 
  # mutate(median_rel_abund = if_else(median_rel_abund == 0, (2/3)*lod, median_rel_abund)) %>%
  ggplot(aes(y = taxon, x = median_rel_abund, color = group)) +
  # geom_vline(xintercept = lod, size = 0.1, linetype = 2) +
  stat_summary(fun.data = median_hilow, geom = "pointrange",
               fun.args = list(conf.int = 0.5),
               position = position_dodge(width = 0.5),
               show.legend = T
               ) +
  coord_trans(x = "log10") +
  scale_x_continuous(limits = c(NA, 100),
                     breaks = c(0.1, 1, 10, 100),
                     labels = c(0.1, 1, 10, 100)) +
labs(y = NULL,
     x = "Relative Abundance (log10)",
     subtitle = "Point Range at 50th Percentile",
     color = NULL) +
  theme_classic() +
  theme_classic() +
  theme(axis.text.y = element_markdown(angle = 0, vjust = 0.5),
        axis.text.x = element_markdown(angle = 0, hjust = 0.5),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        legend.background = element_rect(colour = "gray", fill = NA),
        legend.margin = margin(t = 3, r = 3, b = 3, l = 3))

otuplot
```

# Statistical Tests
> We will use a sub-sample containing same number of OTUs to perform statistical analysis.

## Kruskal Wallis Test
- Kruskal Wallis Test is a non-parametric.
- Assumes no relationship between the members in each group or between groups.

Here we will demonstrate the following:
- Statistical test: Kruskal-Wallis test
- Pairwise test: Paired samples Wilcoxon test or Wilcoxon signed-rank test.
- Correction method: Benjamini-Hochberg (BH).

```{r fig.dim=c(6, 7)}
expt_signif <- genus_median_rel_abund %>% 
  group_by(taxon) %>% 
  nest() %>% 
  mutate(expt_tests = map(.x = data,
                         ~kruskal.test(median_rel_abund ~group, data = .x) %>% 
                           tidy())) %>%
  unnest(expt_tests, names_repair = "unique") %>% 
  ungroup() %>% 
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>% 
  select(taxon, data, p.value, p_adj, method)

if(min(expt_signif$p.value) < 0.2) {
  expt_signif <- expt_signif %>% 
    filter(p.value < 0.2)

print(paste("P-values range from:", min(expt_signif$p.value), "-", max(expt_signif$p.value)))

### Setting quartile limits
get_max_quartile <- function(x) {
  x %>% 
    group_by(group) %>% 
    summarise(third_q = quantile(median_rel_abund, prob = 0.75), .groups = "drop") %>%
    summarise(max_quartile = max(third_q)) %>% 
    pull(max_quartile)
}

### Pairwise comparison
pairs_test <- expt_signif %>% 
  mutate(max_quartile = map_dbl(.x = data, ~get_max_quartile(.x))) %>% 
  mutate(pairwise_tests = map(.x = data,
                              ~pairwise.wilcox.test(x = .x$median_rel_abund,
                                                    g = .x$group,
                                                    p.adjust.method = "BH") %>%
                                tidy())) %>%
  unnest(pairwise_tests, names_repair = "unique") %>% 
  select(taxon, group1, group2, p.value = p.value...3, p_adj, max_quartile)

# Defining x-y and stars coordinates
pairs_test_x_y <- pairs_test %>% 
  select(taxon, group1, group2, p.value, max_quartile) %>% 
  mutate(pos = as.numeric(taxon),
         y = if_else(
           group1 == "Wildebeest", pos + 0.2, pos),
         yend = if_else(group2 == "Buffalo", pos, pos - 0.2),
         x = case_when(
           group1 == "Wildebeest" &
             group2 == "Buffalo" ~ max_quartile * 1.3),
           # group1 == "Wildebeest" &
           #   group2 == "Zebra" ~ max_quartile * 1.6,
           # group1 == "Wildebeest" &
           #   group2 == "Warthog" ~ max_quartile * 1.9),
         xend = x,
         x_star = 1.1*x,
         y_star = case_when(
           group1 == "Wildebeest" &
             group2 == "Buffalo" ~ pos + 0.05))
           # group1 == "Wildebeest" &
           #   group2 == "Zebra" ~ pos + 0.05,
           # group1 == "Wildebeest" &
           #   group2 == "Warthog" ~ pos + 0.05))

pairs_test %>%
  mutate(taxon = str_replace_all(taxon, "\\*", ""),
         taxon = str_replace_all(taxon, "<br>", " ")) %>% 
  saveRDS(., "RDataRDS/pairs_test.rds")

# Add significant stars to the otuplot
print(otuplot +
  geom_segment(data = pairs_test_x_y, aes(x = x, xend = xend, y = y,  yend = yend),
               inherit.aes = F) +
  geom_text(data = pairs_test_x_y,
            aes(x = x_star, y = y_star), label = "*", inherit.aes = F))

# Significant pairs table
print(readRDS("RDataRDS/pairs_test.rds") %>% 
  select(-max_quartile, 
         Taxon = taxon, 
         Group1 = group1, 
         Group2 = group2,
         P_value = p.value,
         P_adj = p_adj,
         ))

# Print error message if not significant.
} else {
  print("No significant values found at `P ≤ 0.05`")
  paste("P-values range from:", min(expt_signif$p.value), "-", max(expt_signif$p.value))
}

```
