# Data Distribution Graphs {#data-distribution}

```{r}
## Load libraries and data
source("R/common.R")
library(tidyverse)
library(ggtext)
library(phyloseq)
library(metagMisc)
library(microbiome)
library(microViz)
library(RColorBrewer)
library(glue)

# View object content
load("../iMAP-part3/RDataRDS/Rjoined_objects.RData", verbose = T)
```

## Barchats

### Stacked sample barcharts filled by taxon

> Note: It is important to get a clear insights into the sample variability before grouping by desired variables.

```{r sample_barchat_fig, fig.height=4, fig.width=6, animation.hook='gifski', interval=4}
load("../iMAP-part3/RDataRDS/Rjoined_objects.RData", verbose = T)

taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "kingdom",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(sample_id, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 2,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = T),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = sample_id, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Kingdom Abundance",
       fill = "Kingdom") +
  theme_classic() +
  theme(axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(8, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip() + nolegend
  

#-------------------
taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "phylum",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
   group_by(sample_id, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 2,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
   group_by(sample_id, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = T),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = sample_id, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Phylum Abundance",
       fill = "Phylum") +
  theme_classic() +
  theme(axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(8, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip() + nolegend

#-------------------
taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "class",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
   group_by(sample_id, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 2,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
   group_by(sample_id, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = T),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = sample_id, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Class Abundance",
       fill = "Class") +
  theme_classic() +
  theme(axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(8, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip() + nolegend

#-------------------
taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "order",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
   group_by(sample_id, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 2,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
   group_by(sample_id, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = T),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = sample_id, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Order Abundance",
       fill = "Order") +
  theme_classic() +
  theme(axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(8, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip() + nolegend

#-------------------
taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "family",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
   group_by(sample_id, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 2,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
   group_by(sample_id, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = T),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = sample_id, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Family Abundance",
       fill = "Family") +
  theme_classic() +
  theme(axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(8, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip() + nolegend

#-------------------
taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "genus",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
   group_by(sample_id, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 2,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
   group_by(sample_id, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = T),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = sample_id, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Genus Abundance",
       fill = "Genus") +
  theme_classic() +
  theme(axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(8, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip() + nolegend
```

<br>

### Stacked group barcharts filled by taxon

```{r group_barchart_fig, fig.height=4, fig.width=6, animation.hook='gifski', interval=4}
load("../iMAP-part3/RDataRDS/Rjoined_objects.RData", verbose = T)

taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "kingdom",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 2,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = T),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = group, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Kingdom Abundance",
       fill = "Kingdom") +
  theme_classic() +
  theme(axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(8, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  centertitle

#-------------------
taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "phylum",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 2,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = T),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = group, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Phylum Abundance",
       fill = "Phylum") +
  theme_classic() +
  theme(axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(8, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  centertitle

#-------------------
taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "class",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 2,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = T),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = group, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Class Abundance",
       fill = "Class") +
  theme_classic() +
  theme(axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(8, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  centertitle

#-------------------
taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "order",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 2,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = T),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = group, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Order Abundance",
       fill = "Order") +
  theme_classic() +
  theme(axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(8, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  centertitle

#-------------------
taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "family",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 2,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = T),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = group, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Family Abundance",
       fill = "Family") +
  theme_classic() +
  theme(axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(8, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  centertitle

#-------------------
taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "genus",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")


taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 2,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = T),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = group, y = median_rel_abund, fill = taxon)) +
  geom_col(position = "fill") +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Genus Abundance",
       fill = "Genus") +
  theme_classic() +
  theme(axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(8, "pt"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  centertitle
```


### Grouped taxa barcharts

```{r taxon_barcharts_fig, fig.height=4, fig.width=6, animation.hook='gifski', interval=4}
set.seed(110912)

load("../iMAP-part3/RDataRDS/Rjoined_objects.RData", verbose = T)

taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "kingdom",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 1,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = taxon, y = median_rel_abund, fill = group)) +
  # geom_col(width = 1, position = position_dodge()) +
  geom_col(width=0.8, position = position_dodge()) +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Kingdom Abundance",
       fill = "Group") +
  theme_classic() +
  theme(axis.text.x = element_markdown(angle = 0, hjust = 1, vjust = 1),
        axis.text.y = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        panel.background = element_blank(),
        panel.grid.major.x =  element_line(colour = "lightgray", size = 0.1),
        panel.border = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  centertitle + 
  coord_flip() + ybreaks5


taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "phylum",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 1,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = taxon, y = median_rel_abund, fill = group)) +
  # geom_col(width = 1, position = position_dodge()) +
  geom_col(width=0.8, position = position_dodge()) +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Phylum Abundance",
       fill = "Group") +
  theme_classic() +
  theme(axis.text.x = element_markdown(angle = 0, hjust = 1, vjust = 1),
        axis.text.y = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        panel.background = element_blank(),
        panel.grid.major.x =  element_line(colour = "lightgray", size = 0.1),
        panel.border = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  centertitle + 
  coord_flip() + ybreaks5


taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "class",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 1,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = taxon, y = median_rel_abund, fill = group)) +
  # geom_col(width = 1, position = position_dodge()) +
  geom_col(width=0.8, position = position_dodge()) +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Class Abundance",
       fill = "Group") +
  theme_classic() +
  theme(axis.text.x = element_markdown(angle = 0, hjust = 1, vjust = 1),
        axis.text.y = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        panel.background = element_blank(),
        panel.grid.major.x =  element_line(colour = "lightgray", size = 0.1),
        panel.border = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  centertitle + 
  coord_flip() + ybreaks5


taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "order",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 1,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = taxon, y = median_rel_abund, fill = group)) +
  # geom_col(width = 1, position = position_dodge()) +
  geom_col(width=0.8, position = position_dodge()) +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Order Abundance",
       fill = "Group") +
  theme_classic() +
  theme(axis.text.x = element_markdown(angle = 0, hjust = 1, vjust = 1),
        axis.text.y = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        panel.background = element_blank(),
        panel.grid.major.x =  element_line(colour = "lightgray", size = 0.1),
        panel.border = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  centertitle + 
  coord_flip() + ybreaks5


taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "family",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 1,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = taxon, y = median_rel_abund, fill = group)) +
  # geom_col(width = 1, position = position_dodge()) +
  geom_col(width=0.8, position = position_dodge()) +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Family Abundance",
       fill = "Group") +
  theme_classic() +
  theme(axis.text.x = element_markdown(angle = 0, hjust = 1, vjust = 1),
        axis.text.y = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        panel.background = element_blank(),
        panel.grid.major.x =  element_line(colour = "lightgray", size = 0.1),
        panel.border = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  centertitle + 
  coord_flip() + ybreaks5


taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "genus",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = sum(rel_abund), .groups = "drop") %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = 100*median(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(taxon) %>%
  summarise(pool = max(median_rel_abund) < 1,
            median = median(median_rel_abund),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(group, taxon) %>% 
  summarise(median_rel_abund = sum(median_rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = taxon, y = median_rel_abund, fill = group)) +
  # geom_col(width = 1, position = position_dodge()) +
  geom_col(width=0.8, position = position_dodge()) +
labs(x = NULL,
       y = "Median Relative Abundance (%)",
       title = "Genus Abundance",
       fill = "Group") +
  theme_classic() +
  theme(axis.text.x = element_markdown(angle = 0, hjust = 1, vjust = 1),
        axis.text.y = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        panel.background = element_blank(),
        panel.grid.major.x =  element_line(colour = "lightgray", size = 0.1),
        panel.border = element_blank()) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous(expand = c(0, 0)) +
  centertitle + 
  coord_flip() 
```


## Strip chart

```{r strip_fig, fig.height=4, fig.width=6, animation.hook='gifski', interval=4}
set.seed(110912)

load("../iMAP-part3/RDataRDS/Rjoined_objects.RData", verbose = T)

taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "genus",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = 100*sum(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(group, taxon) %>%
  summarise(median = median(rel_abund), .groups = "drop") %>%
  group_by(taxon) %>% 
  summarise(pool = max(median) < 3,
            median = median(median),
            .groups = "drop")


inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, group, taxon) %>% 
  summarise(rel_abund = sum(rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = T),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = taxon, y = rel_abund, fill = group)) +
  geom_jitter(position = position_jitterdodge(
    jitter.width = 0.3, 
    dodge.width = 0.4),
    pch = 21, 
    stroke = 0,
    size = 1.8) +
  # stat_summary(fun=median, geom = "crossbar",
  #              position = position_dodge(width = 0.8),
  #              width = 0.8, 
  #              size = 0.25,
  #              show.legend = F) +
  scale_color_manual(name = NULL,
                    breaks = c("Buffalo",
                               "Warthog",
                               "Wildebeest",
                               "Zebra"),
                    labels = c("Buffalo",
                               "Warthog",
                               "Wildebeest",
                               "Zebra"),                   
                    values = c("blue",
                               "red",
                               "green",
                               "purple")) +
labs(x = NULL,
       y = "Median Relative Abundance (%)") +
  theme_classic() +
  theme(axis.text.x = element_markdown(angle = 90, hjust = 1, vjust = 1),
        axis.text.y = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        legend.background = element_rect(colour = "gray", fill = NA),
        legend.margin = margin(t = 5, r = 5, b = 5, l = 5)) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous() +
  centertitle + 
  coord_cartesian(ylim = c(0, 100)) #+ 
  # coord_flip() 
 
```



## Histogram with bins

```{r histogram_fig, fig.height=4, fig.width=6, animation.hook='gifski', interval=4}
### Standard boxplot

set.seed(110912)

load("../iMAP-part3/RDataRDS/Rjoined_objects.RData", verbose = T)

taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "genus",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = 100*sum(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(group, taxon) %>%
  summarise(median = median(rel_abund), .groups = "drop") %>%
  group_by(taxon) %>% 
  summarise(pool = max(median) < 3,
            median = median(median),
            .groups = "drop")


inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, group, taxon) %>% 
  summarise(rel_abund = sum(rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = T),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = taxon, y = rel_abund)) +
  geom_histogram(stat = "identity", bins = 20) +

  # stat_summary(fun=median, geom = "crossbar",
  #              position = position_dodge(width = 0.8),
  #              width = 0.8, 
  #              size = 0.25,
  #              show.legend = F) +
  scale_color_manual(name = NULL,
                    breaks = c("Buffalo",
                               "Warthog",
                               "Wildebeest",
                               "Zebra"),
                    labels = c("Buffalo",
                               "Warthog",
                               "Wildebeest",
                               "Zebra"),                   
                    values = c("blue",
                               "red",
                               "green",
                               "purple")) +
labs(y = NULL,
       x = "Relative Abundance (%)",
       title = "Standard box-jitter plot") +
  theme_classic() +
  theme(axis.text.x = element_markdown(angle = 90, hjust = 1, vjust = 1),
        axis.text.y = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        legend.background = element_rect(colour = "gray", fill = NA),
        legend.margin = margin(t = 5, r = 5, b = 5, l = 5)) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous() +
  centertitle + 
  coord_cartesian(ylim = c(0, 100))

```


## Jitter and Boxplot

```{r jitter_box_fig, fig.height=4, fig.width=6, animation.hook='gifski', interval=4}
### Standard boxplot

set.seed(110912)

load("../iMAP-part3/RDataRDS/Rjoined_objects.RData", verbose = T)

taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "genus",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = 100*sum(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(group, taxon) %>%
  summarise(median = median(rel_abund), .groups = "drop") %>%
  group_by(taxon) %>% 
  summarise(pool = max(median) < 3,
            median = median(median),
            .groups = "drop")


inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, group, taxon) %>% 
  summarise(rel_abund = sum(rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = T),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(x = taxon, y = rel_abund, color = group)) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.8)) +
  geom_jitter(position = position_jitterdodge(
    jitter.width = 0.3,
    dodge.width = 0.4),
    pch = 19,
    stroke = 0,
    size = 1.8) +
  # stat_summary(fun=median, geom = "crossbar",
  #              position = position_dodge(width = 0.8),
  #              width = 0.8, 
  #              size = 0.25,
  #              show.legend = F) +
  scale_color_manual(name = NULL,
                    breaks = c("Buffalo",
                               "Warthog",
                               "Wildebeest",
                               "Zebra"),
                    labels = c("Buffalo",
                               "Warthog",
                               "Wildebeest",
                               "Zebra"),                   
                    values = c("blue",
                               "red",
                               "green",
                               "purple")) +
labs(y = NULL,
       x = "Relative Abundance (%)",
       title = "Standard box-jitter plot") +
  theme(axis.text.x = element_markdown(angle = 90, hjust = 1, vjust = 1),
        axis.text.y = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        legend.background = element_rect(colour = "gray", fill = NA),
        legend.margin = margin(t = 5, r = 5, b = 5, l = 5)) +
  guides(fill = guide_legend(ncol=1)) +
  scale_y_continuous() +
  centertitle + 
  coord_cartesian(ylim = c(0, 100))


### Flip x-y coordinates and add statistics
set.seed(110912)

taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "genus",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = 100*sum(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(group, taxon) %>%
  summarise(median = median(rel_abund), .groups = "drop") %>%
  group_by(taxon) %>% 
  summarise(pool = max(median) < 3,
            median = median(median),
            .groups = "drop")


inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, group, taxon) %>% 
  summarise(rel_abund = sum(rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(y = taxon, x = rel_abund, color = group)) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.8)) +
  geom_jitter(position = position_jitterdodge(
    jitter.width = 0.3,
    dodge.width = 0.4),
    pch = 19) +
  # stat_summary(fun = median, geom = "crossbar",
  #              position = position_dodge(width = 0.8),
  #              width = 0.8,
  #              size = 0.25,
  #              show.legend = F) +
  scale_color_manual(name = NULL,
                    breaks = c("Buffalo",
                               "Warthog",
                               "Wildebeest",
                               "Zebra"),
                    labels = c("Buffalo",
                               "Warthog",
                               "Wildebeest",
                               "Zebra"),                   
                    values = c("blue",
                               "red",
                               "green",
                               "purple")) +
labs(y = NULL,
       x = "Relative Abundance (%)",
       title = "Flipped box-jitter plot") +
  theme_classic() +
  theme(axis.text.y = element_markdown(angle = 0, hjust = 1, vjust = 1),
        # axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        legend.background = element_rect(colour = "gray", fill = NA),
        legend.margin = margin(t = 5, r = 5, b = 5, l = 5)) +
  guides(fill = guide_legend(ncol=1)) +
  scale_x_continuous()
 
### Jitter plot with statistics

set.seed(110912)

taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "genus",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = 100*sum(rel_abund), .groups = "drop")

taxon_pool <- taxon_rel_abund %>%
  group_by(group, taxon) %>%
  summarise(median = median(rel_abund), .groups = "drop") %>%
  group_by(taxon) %>% 
  summarise(pool = max(median) < 3,
            median = median(median),
            .groups = "drop")


inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, group, taxon) %>% 
  summarise(rel_abund = sum(rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(y = taxon, x = rel_abund, color = group)) +
  # geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.8)) +
  geom_jitter(position = position_jitterdodge(
    jitter.width = 0.5,
    dodge.width = 0.4),
    pch = 19) +
  stat_summary(fun.data = median_hilow, geom = "crossbar",
               fun.args = list(conf.int = 0.5),
               position = position_dodge(width = 0.8),
               width = 0.8,
               size = 0.25,
               show.legend = F) +
  scale_color_manual(name = NULL,
                    breaks = c("Buffalo",
                               "Warthog",
                               "Wildebeest",
                               "Zebra"),
                    labels = c("Buffalo",
                               "Warthog",
                               "Wildebeest",
                               "Zebra"),                   
                    values = c("blue",
                               "red",
                               "green",
                               "purple")) +
labs(y = NULL,
       x = "Relative Abundance (%)",
       title = "Flipped jitter plot with statistics") +
  theme_classic() +
  theme(axis.text.y = element_markdown(angle = 0, hjust = 1, vjust = 1),
        # axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        legend.background = element_rect(colour = "gray", fill = NA),
        legend.margin = margin(t = 5, r = 5, b = 5, l = 5)) +
  guides(fill = guide_legend(ncol=1)) +
  scale_x_continuous()


 inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, group, taxon) %>% 
  summarise(rel_abund = sum(rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0)) %>%
  ggplot(aes(y = taxon, x = rel_abund, color = group)) +
  # geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.8)) +
  # geom_jitter(position = position_jitterdodge(
  #   jitter.width = 0.3,
  #   dodge.width = 0.4),
  #   pch = 19) +
  stat_summary(fun.data = median_hilow, geom = "pointrange",
               fun.args = list(conf.int = 0.5),
               position = position_dodge(width = 0.4),
               show.legend = T) +
  scale_color_manual(name = NULL,
                    breaks = c("Buffalo",
                               "Warthog",
                               "Wildebeest",
                               "Zebra"),
                    # labels = c("A",
                    #            "B",
                    #            "C",
                    #            "D"),                    
                    values = c("blue",
                               "red",
                               "green",
                               "purple")) +
labs(y = NULL,
       x = "Relative Abundance (%)",
       title = "Point Range at 50th Percentile") +
  theme_classic() +
  theme(axis.text.y = element_markdown(angle = 0, hjust = 1, vjust = 1),
        # axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        legend.background = element_rect(colour = "gray", fill = NA),
        legend.margin = margin(t = 3, r = 3, b = 3, l = 0))
  guides(fill = guide_legend(ncol=1)) +
  scale_x_continuous()
 
```

## Log scaled axis
> Note: Log transformation may introduce infinite values in continuous x-axis for zero relative abundance. The rows containing non-finite values will be removed. Solution is replace the zeros with a tiny non-significant value below the LoD (limit od detection).

Let's establish the `lod`
```{r}
set.seed(110912)

nseqs_per_sample <- readRDS("../iMAP-part3/RDataRDS/mo_otutable_long.rds") %>% 
  group_by(sample_id) %>%
  summarise(N = sum(count), .groups="drop") %>%
  count(N) %>%
  pull(N)

# stopifnot(length(nseqs_per_sample) == 1)
stopifnot(length(nseqs_per_sample) != 1)

lod <- max(100* 1/nseqs_per_sample)
lod
```

```{r fig.dim=c(7, 6)}
load("../iMAP-part3/RDataRDS/Rjoined_objects.RData", verbose = T)

taxon_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "genus",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)
         ) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = 100*sum(rel_abund), .groups = "drop") %>% 
  mutate(
    taxon = str_replace(taxon,
                             "(.*)_unclassified", "Unclassified*\\1*"),
    taxon = str_replace(taxon,
                             "^([^>]*)$", "*\\1*"),
    taxon = str_replace_all(taxon,
                             "_", " "))

taxon_pool <- taxon_rel_abund %>%
  group_by(group, taxon) %>%
  summarise(median = median(rel_abund), .groups = "drop") %>%
  group_by(taxon) %>% 
  summarise(pool = max(median) < 1,
            median = median(median),
            .groups = "drop")

inner_join(taxon_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, group, taxon) %>% 
  summarise(rel_abund = sum(rel_abund), 
            median = min(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0)) %>%
  mutate(rel_abund = if_else(rel_abund == 0, (2/3)*lod, rel_abund)) %>% 
  ggplot(aes(y = taxon, x = rel_abund, color = group)) +
  geom_vline(xintercept = lod, size = 0.1, linetype = 2) +
 
  stat_summary(fun.data = median_hilow, geom = "pointrange",
               fun.args = list(conf.int = 0.5),
               position = position_dodge(width = 0.6),
               show.legend = T
               ) +
  coord_trans(x = "log10") +
  scale_x_continuous(limits = c(NA, 100),
                     breaks = c(0.1, 1, 10, 100),
                     labels = c(0.1, 1, 10, 100)) +
  scale_color_manual(name = NULL,
                    breaks = c("Buffalo",
                               "Warthog",
                               "Wildebeest",
                               "Zebra"),
                    labels = c("Buffalo",
                               "Warthog",
                               "Wildebeest",
                               "Zebra"),                   
                    values = c("blue",
                               "red",
                               "green",
                               "purple")) +
labs(y = NULL,
       x = "Relative Abundance (log10)",
       title = "Point Range at 50th Percentile") +
  theme_classic() +
  theme(axis.text.y = element_markdown(angle = 0, hjust = 1, vjust = 1),
        # axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        legend.background = element_rect(colour = "gray", fill = NA),
        legend.margin = margin(t = 3, r = 0, b = 3, l = 0))
```

## Gluing OTU ID to taxon
```{r fig.dim=c(7, 6)}
load("../iMAP-part3/RDataRDS/Rjoined_objects.RData", verbose = T)

otu_rel_abund <- mo_Rjoined_object %>% 
  rename(group = isolate) %>% 
  filter(level == "taxon_phylo",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*",taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = 100*sum(rel_abund), .groups = "drop") %>% 
    mutate(
    taxon = str_replace(taxon,
                             "(.*)_unclassified", "Unclassified*\\1*"),
    taxon = str_replace(taxon,
                             "^([^>]*)$", "*\\1*"),
    taxon = str_replace_all(taxon,
                             "_", " "))

taxon_pool <- otu_rel_abund %>%
  group_by(group, taxon) %>%
  summarise(median = median(rel_abund), .groups = "drop") %>%
  group_by(taxon) %>% 
  summarise(pool = max(median) < 1,
            median = max(median),
            .groups = "drop")

inner_join(otu_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, group, taxon) %>% 
  summarise(rel_abund = sum(rel_abund), 
            median = max(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0)) %>%
  mutate(rel_abund = if_else(rel_abund == 0, 2/3*lod, rel_abund)) %>% 
  ggplot(aes(y = taxon, x = rel_abund, color = group)) +
  geom_vline(xintercept = lod, size = 0.1, linetype = 2) +
  stat_summary(fun.data = median_hilow, geom = "pointrange",
               fun.args = list(conf.int = 0.5),
               position = position_dodge(width = 0.6),
               # show.legend = T
               ) +
  coord_trans(x = "log10") +
  scale_x_continuous(limits = c(NA, 100),
                     breaks = c(0.1, 1, 10, 100),
                     labels = c(0.1, 1, 10, 100)) +
  scale_color_manual(name = NULL,
                    breaks = c("Buffalo",
                               "Warthog",
                               "Wildebeest",
                               "Zebra"),
                    labels = c("Buffalo",
                               "Warthog",
                               "Wildebeest",
                               "Zebra"),                   
                    values = c("blue",
                               "red",
                               "green",
                               "purple")) +
labs(y = NULL,
       x = "Relative Abundance (log10)",
       title = "Point Range at 50th Percentile") +
  theme_classic() +
  theme(axis.text.y = element_markdown(angle = 0, hjust = 1, vjust = 1),
        axis.text.x = element_markdown(),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        legend.background = element_rect(colour = "gray", fill = NA),
        legend.margin = margin(t = 3, r = 0, b = 3, l = 0))
```


