# (PART) STATISTICAL ANALYSIS {-}

# Statistical Tests
> We will use a sub-sample containing same number of OTUs to perform statistical analysis.

## Kruskal Wallis Test
- Kruskal Wallis Test is a non-parametric.
- Assumes no relationship between the members in each group or between groups.

Here we will demonstrate the following:
- Statistical test: Kruskal-Wallis test
- Pairwise test: Paired samples Wilcoxon test or Wilcoxon signed-rank test.
- Correction method: Benjamini-Hochberg (BH).

## Preparing data for statistical analysis {#stats-data}
```{r}
source("R/common.R")
library(tidyverse)
library(ggtext)
library(glue)
library(broom)

set.seed(110912)

load("../iMAP-part3/RDataRDS/Rjoined_objects.RData", verbose = T)
```

## Group relative abundance plot
```{r fig.height=8, fig.width=6}
set.seed(110912)

otu_rel_abund <- Rjoined_sub_object %>% 
  select(-taxon) %>% 
  rename(taxon = taxon_phylo) %>% 
  filter(level == "genus",
         group == "Buffalo" | group == "Wildebeest",
         !grepl(".*unassigned.*|.*nclassified.*|.*ncultured.*", taxon)) %>% 
  group_by(sample_id, taxon, group) %>% 
  summarise(rel_abund = 100*sum(rel_abund), .groups = "drop")

taxon_pool <- otu_rel_abund %>%
  group_by(group, taxon) %>%
  summarise(median = median(rel_abund), .groups = "drop") %>%
  group_by(taxon) %>% 
  summarise(pool = max(median) < 3,
            median = max(median),
            .groups = "drop")

## Group relative abundance
otu_group_rel_abund <- inner_join(otu_rel_abund, taxon_pool, by = "taxon" ) %>% 
  mutate(taxon = if_else(pool, "Other", as.character(taxon))) %>% 
  group_by(sample_id, group, taxon) %>% 
  summarise(rel_abund = sum(rel_abund), 
            median = max(median),
            .groups = "drop") %>%
  mutate(taxon = factor(taxon),
         taxon = fct_reorder(taxon, median, .desc = F),
         taxon = fct_shift(taxon, n = 0)) 

## Plot group relative abundance
otuplot <- genus_median_rel_abund %>% 
  mutate(rel_abund = if_else(rel_abund == 0, (2/3)*lod, rel_abund)) %>%
  ggplot(aes(y = taxon, x = rel_abund, color = group)) +
  geom_vline(xintercept = lod, size = 0.1, linetype = 2) +
  stat_summary(fun.data = median_hilow, geom = "pointrange",
               fun.args = list(conf.int = 0.2),
               position = position_dodge(width = 0.5),
               show.legend = T
               ) +
  coord_trans(x = "log10") +
  scale_x_continuous(limits = c(NA, 100),
                     breaks = c(0.1, 1, 10, 100),
                     labels = c(0.1, 1, 10, 100)) +
  scale_color_manual(name = NULL,
                    breaks = c("Wildebeest", "Buffalo", "Zebra", "Warthog"),
                    labels = c("Wildebeest", "Buffalo", "Zebra", "Warthog"),
                    values = c("blue1", "red", "green2", "purple")) +
labs(y = NULL,
       x = "Relative Abundance (log10)",
       title = "Group Point Range at 50th Percentile") +
  theme_classic() +
  theme(axis.text.y = element_markdown(angle = 0, vjust = 0.5),
        axis.text.x = element_markdown(angle = 0, hjust = 0.5),
        legend.text = element_markdown(),
        legend.key.size = unit(12, "pt"),
        legend.background = element_rect(colour = "gray", fill = NA),
        legend.margin = margin(t = 3, r = 3, b = 3, l = 3))

otuplot
```


## Genarate testing data
```{r fig.height=8, fig.width=6}
expt_signif <- genus_median_rel_abund %>% 
  group_by(taxon) %>% 
  nest() %>% 
  mutate(expt_tests = map(.x = data,
                         ~kruskal.test(median_rel_abund ~group, data = .x) %>% 
                           tidy())) %>%
  unnest(expt_tests, names_repair = "unique") %>% 
  ungroup() %>% 
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>% 
  select(taxon, data, p.value, p_adj, method)

if(min(expt_signif$p.value) < 0.05) {
  expt_signif <- expt_signif %>% 
    filter(p.value < 0.05)

### Setting quartile limits
get_max_quartile <- function(x) {
  x %>% 
    group_by(group) %>% 
    summarise(third_q = quantile(rel_abund, prob = 0.75), .groups = "drop") %>%
    summarise(max_quartile = max(third_q)) %>% 
    pull(max_quartile)
}

### Pairwise comparison
pairs_test <- expt_signif %>% 
  mutate(max_quartile = map_dbl(.x = data, ~get_max_quartile(.x))) %>% 
  mutate(pairwise_tests = map(.x = data,
                              ~pairwise.wilcox.test(x = .x$rel_abund,
                                                    g = .x$group,
                                                    p.adjust.method = "BH") %>%
                                tidy())) %>%
  unnest(pairwise_tests, names_repair = "unique") %>% 
  select(taxon, group1, group2, p.value = p.value...3, p_adj, max_quartile)

# Defining x-y and stars coordinates
pairs_test_x_y <- pairs_test %>% 
  select(taxon, group1, group2, p.value, max_quartile) %>% 
  mutate(pos = as.numeric(taxon),
         y = if_else(
           group1 == "Wildebeest", pos + 0.2, pos),
         yend = if_else(group2 == "Buffalo", pos, pos - 0.2),
         x = case_when(
           group1 == "Wildebeest" &
             group2 == "Buffalo" ~ max_quartile * 1.3),
           # group1 == "Wildebeest" &
           #   group2 == "Zebra" ~ max_quartile * 1.6,
           # group1 == "Wildebeest" &
           #   group2 == "Warthog" ~ max_quartile * 1.9),
         xend = x,
         x_star = 1.07*x,
         y_star = case_when(
           group1 == "Wildebeest" &
             group2 == "Buffalo" ~ pos + 0.05))
           # group1 == "Wildebeest" &
           #   group2 == "Zebra" ~ pos + 0.05,
           # group1 == "Wildebeest" &
           #   group2 == "Warthog" ~ pos + 0.05))

pairs_test %>%
  mutate(taxon = str_replace_all(taxon, "\\*", ""),
         taxon = str_replace_all(taxon, "<br>", " ")) %>% 
  saveRDS(., "RDataRDS/pairs_test.rds")

# Add significant stars to the otuplot
print(otuplot +
  geom_segment(data = pairs_test_x_y, aes(x = x, xend = xend, y = y,  yend = yend),
               inherit.aes = F) +
  geom_text(data = pairs_test_x_y,
            aes(x = x_star, y = y_star), label = "*", inherit.aes = F))

# Significant pairs table
print(readRDS("RDataRDS/pairs_test.rds") %>% 
  select(-max_quartile, 
         Taxon = taxon, 
         Group1 = group1, 
         Group2 = group2,
         P_value = p.value,
         P_adj = p_adj,
         ))

# Print error message if not significant.
} else {
  print("No significant values found at `P ≤ 0.05`")
  paste("P-values range from:", min(expt_signif$p.value), "-", max(expt_signif$p.value))
}

```



