# (PART) MICROBIAL DIVERSITY {-}
# Alpha Diversity Analysis {#alpha-diversity}

```{r include=FALSE}
library(vegan)
library(tidyverse)
library(phyloseq)
knitr::opts_chunk$set(echo = F)
path_shared <- "~/Dropbox/CDILLC/GIT_REPOS/smda-end2end/data/final.tx.shared"
```

## Basic diversity metrics {#count_n_metrics}
```{r}
set.seed(2022)

shared <- read_tsv(path_shared, show_col_types = F) %>%
select(sample_id = Group, starts_with("Phylo")) %>% 
  pivot_longer(-sample_id, names_to = "otu", values_to = "count") %>%
  group_by(sample_id) %>%
  mutate(total = sum(count)) %>%
  filter(total > 0) %>%
  group_by(otu) %>%
  mutate(total = sum(count)) %>%
  filter(total != 0) %>%
  ungroup() %>%
  select(-total)

count <- shared %>% 
  group_by(sample_id) %>% 
  summarise(count = sum(count), .groups = "drop")

## Random sampling
rand <- shared %>%
  uncount(count) %>%
  mutate(otu = sample(otu)) %>%
  count(sample_id, otu, name = "count")

## Diversity metrics
source("R/alpha.R")
set.seed(2022)

div_metrics <- shared %>%
  group_by(sample_id) %>%
  summarise(sobs = richness(count),
            shannon = shannon(count),
            simpson = simpson(count),
            invsimpson = 1/simpson) %>% 
  inner_join(count, ., by = "sample_id")

div_metrics %>% 
  saveRDS("RDataRDS/div_metrics.rds")

div_metrics %>% 
  kableExtra::kable() %>% 
  kableExtra::kable_styling(full_width = F, bootstrap_options = "basic")
```


## Diversity plots

### Using original count
```{r message=FALSE, warning=FALSE}
readRDS("RDataRDS/div_metrics.rds") %>%
  pivot_longer(cols=c(sobs, shannon, invsimpson, simpson),
               names_to="metric") %>%
  ggplot(aes(x=count, y=value)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~metric, nrow=4, scales="free_y") + 
  theme_bw() +
  theme(panel.grid = element_blank())
```

### Using randomly sampled count
```{r}
source("R/alpha.R")
library(vegan)
set.seed(2022)

rand %>%
  group_by(sample_id) %>%
  summarise(sobs = richness(count),
            shannon = shannon(count),
            simpson = simpson(count),
            invsimpson = 1/simpson,
            n = sum(count)) %>%
  pivot_longer(cols=c(sobs, shannon, invsimpson, simpson),
               names_to="metric") %>%
  ggplot(aes(x=n, y=value)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~metric, nrow=4, scales="free_y") + 
  theme_bw() +
  theme(panel.grid = element_blank())
```


### Using phyloseq object
> Note: Additional metrics are computed automatically when using phyloseq object
```{r fig.dim=c(8, 10)}
library(phyloseq)
library(tidyverse)
load("../iMAP-part3/RDataRDS/phyloseq_objects.RData", verbose = F)
ps <- ps_raw

set.seed(1234)

alpha_div_tbl <- data.frame(
  "sample_id" = sample_names(ps),
  "group" = phyloseq::sample_data(ps)$isolate,
  "n_read" = phyloseq::sample_sums(ps),
  phyloseq::estimate_richness(ps),
  "phylodiv" = picante::pd(samp = data.frame(t(data.frame(phyloseq::otu_table(ps)))), 
                           tree = phyloseq::phy_tree(ps))[, 1])

saveRDS(alpha_div_tbl, file = "RDataRDS/ps_alpha_div_tbl.rds")

readRDS("RDataRDS/ps_alpha_div_tbl.rds") %>%
  rename(sobs = Observed, 
         chao = Chao1, 
         ace = ACE,
         simpson = Simpson, 
         invsimpson = InvSimpson, 
         shannon = Shannon,
         fisher = Fisher) %>% 
  pivot_longer(cols=c(sobs, 
                      chao, 
                      ace,
                      simpson, 
                      invsimpson, 
                      shannon,
                      fisher,
                      phylodiv),
               names_to="metric", 
               values_to = "value") %>% 
  ggplot(aes(x = n_read, y=value)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~metric, nrow=8, scales="free_y") + 
  theme_bw() +
  theme(panel.grid = element_blank())
```

## Sampling coverage 
```{r}
sampling_coverage <- shared %>% 
  group_by(sample_id) %>% 
summarise(n_seq = sum(count)) %>% 
arrange(n_seq)
```

### Histogram
```{r}
sampling_coverage %>%
  ggplot(aes(x = n_seq)) +
  geom_histogram()
```


### Line plot
```{r}
set.seed(2022)
sampling_coverage %>%
  arrange(n_seq) %>%
  ggplot(aes(x = 1:nrow(.), y= n_seq)) +
  geom_line() +
  scale_x_continuous(n.breaks = nrow(sampling_coverage)) +
  labs(x = "Data Point", y = " Number Sequences") +
  theme_bw()
```

### Point plot
```{r}
set.seed(2022)
sampling_coverage %>%
  arrange(n_seq) %>%
  ggplot(aes(x = 1:nrow(.), y= n_seq)) +
  geom_jitter() +
  scale_x_continuous(n.breaks = nrow(sampling_coverage)) +
  labs(x = "Data Point", y = " Number Sequences") +
  theme_bw()
```

### Singletons and coverage statistics
```{r}
set.seed(2022)
coverage_stats <- shared %>%
  group_by(sample_id) %>%
  summarise(n_seqs = sum(count),
            n_sings =sum(count == 1),
            coverage = 100*(1 - n_sings/n_seqs)) %>%
  filter(n_seqs > 0) %>%
  as.data.frame() %>%
  write_tsv("RDataRDS/sample_coverage.tsv")

as.data.frame(read_tsv("RDataRDS/sample_coverage.tsv", show_col_types = F)) %>% 
  kableExtra::kable(format = "html") %>% 
  kableExtra::kable_styling(full_width = F, bootstrap_options = "basic")
```


<!-- # Hypothesis Testing for Alpha Diversity -->
<!-- - Assess the variation in alpha-diversity between groups. -->
<!-- - Then accept or reject the null hypothesis that there is no difference in location between groups. -->
<!-- - Box plots can show the location of the median. -->
<!-- - Perform samples Wilcoxon test. -->

<!-- ## Data distribution -->
<!-- ```{r} -->
<!-- count_tbl %>% -->
<!--   gather(key = metric, value = value, c("Observed", "Shannon", "PhyloDiv")) %>% -->
<!--   mutate(metric = factor(metric, levels = c("Observed", "Shannon", "PhyloDiv"))) %>% -->
<!--   ggplot(aes(x = Group, y = value)) + -->
<!--   geom_boxplot(outlier.color = NA) + -->
<!--   geom_jitter(aes(color = Group), height = 0, width = .2) + -->
<!--   labs(x = "", y = "") + -->
<!--   facet_wrap(~ metric, scales = "free") + -->
<!--   theme(legend.position="none") -->

<!-- ``` -->

<!-- ## Summarize -->
<!-- ```{r} -->
<!-- count_tbl %>% -->
<!--   group_by(Group) %>% -->
<!--     dplyr::summarize( -->
<!--       median_observed = median(Observed), -->
<!--       median_invsimpson = median(InvSimpson), -->
<!--       median_shannon = median(Shannon), -->
<!--       median_pd = median(PhyloDiv)) -->

<!-- ``` -->

<!-- ## Paired Wilcoxon test -->

<!-- ### Observed vs Group -->
<!-- ```{r} -->
<!-- wilcox.test(Observed ~ Group, data = count_tbl, exact = FALSE, conf.int = TRUE) -->

<!-- ``` -->

<!-- ### Shannon vs Group -->
<!-- ```{r} -->
<!-- wilcox.test(Shannon ~ Group, data = count_tbl, exact = FALSE, conf.int = TRUE) -->

<!-- ``` -->

<!-- ### InvSimpson vs Group -->
<!-- ```{r} -->
<!-- wilcox.test(InvSimpson ~ Group, data = count_tbl, exact = FALSE, conf.int = TRUE) -->

<!-- ``` -->

<!-- ### Phylo Diversity vs Group -->
<!-- ```{r} -->
<!-- wilcox.test(PhyloDiv ~ Group, data = count_tbl, exact = FALSE, conf.int = TRUE) -->

<!-- ``` -->


<!-- ### Coverage line plot -->
<!-- ```{r} -->
<!-- set.seed(2022) -->
<!-- sampling_coverage %>%  -->
<!--   arrange(n_seq) %>%  -->
<!--   ggplot(aes(x = 1:nrow(.), y= n_seq)) + -->
<!--   geom_line() + -->
<!--   coord_cartesian(xlim = c(0, nrow(sampling_coverage))) -->
<!-- ``` -->

<!-- ### Coverage log-scaled jitter -->
<!-- ```{r} -->
<!-- set.seed(2022) -->
<!-- sampling_coverage %>%  -->
<!--   arrange(n_seq) %>%  -->
<!--   ggplot(aes(x = 1:nrow(.), y= n_seq)) + -->
<!--   geom_jitter() + -->
<!--   scale_y_log10() -->
<!-- ``` -->

<!-- ### Coverage statistics -->
<!-- ```{r} -->
<!-- set.seed(2022) -->
<!-- coverage_stats <- shared %>%  -->
<!--   group_by(sample_id) %>%  -->
<!--   summarise(n_seqs = sum(count), -->
<!--             n_sings =sum(count == 1), -->
<!--             coverage = 100*(1 - n_sings/n_seqs)) %>%  -->
<!--   filter(n_seqs > 0) %>%  -->
<!--   as.data.frame() %>%  -->
<!--   write_tsv("RDataRDS/sample_coverage.tsv") -->

<!-- as.data.frame(read_tsv("RDataRDS/sample_coverage.tsv", show_col_types = F)) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- set.seed(2022) -->
<!-- coverage_stats %>%  -->
<!--   arrange(-coverage) %>%  -->
<!--   ggplot(aes(x = n_seqs, y = coverage)) + -->
<!--                geom_point() -->
<!-- ``` -->
