# (PART) STATISTICAL ANALYSIS {-}

# Statistical Tests
> We will use a sub-sample to perform statistical analysis.

## Load libraries and data
```{r}
source("R/common.R")
source("R/lefse.R")
library(ggtext)
library(RColorBrewer)
library(glue)
library(broom)
library(tidyverse)

load("RDataRDS/taxa_abund_data.RData", verbose = T)
```

## Kruskal Wallis Test
- Kruskal Wallis Test is a non-parametric.
- Assumes no relationship between the members in each group or between groups.

Here we will demonstrate the following:
- Statistical test: Kruskal-Wallis test
- Pairwise test: Paired samples Wilcoxon test or Wilcoxon signed-rank test.
- Correction method: Benjamini-Hochberg (BH).

```{r kruskal_wallis_fig, fig.dim=c(6, 7)}
expt_signif <- genus_median_rel_abund %>% 
  group_by(taxon) %>% 
  nest() %>% 
  mutate(expt_tests = map(.x = data,
                         ~kruskal.test(median_rel_abund ~group, data = .x) %>% 
                           tidy())) %>%
  unnest(expt_tests, names_repair = "unique") %>% 
  ungroup() %>% 
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>% 
  dplyr::select(taxon, data, p.value, p_adj, method)
if(min(expt_signif$p.value) < 0.2) {
  expt_signif <- expt_signif %>% 
    filter(p.value < 0.2)
cat(paste("The computed P-values for this test range from:\n"),
paste(min(expt_signif$p.value), "-",
      max(expt_signif$p.value)))
### Setting quartile limits
get_max_quartile <- function(x) {
  x %>% 
    group_by(group) %>% 
    summarise(third_q = quantile(median_rel_abund, prob = 0.75), .groups = "drop") %>%
    summarise(max_quartile = max(third_q)) %>% 
    pull(max_quartile)
}
### Pairwise comparison
pairs_test <- expt_signif %>% 
  mutate(max_quartile = map_dbl(.x = data, ~get_max_quartile(.x))) %>% 
  mutate(pairwise_tests = map(.x = data,
                              ~pairwise.wilcox.test(x = .x$median_rel_abund,
                                                    g = .x$group,
                                                    p.adjust.method = "BH") %>%
                                tidy())) %>%
  unnest(pairwise_tests, names_repair = "unique") %>% 
  dplyr::select(taxon, group1, group2, p.value = p.value...3, p_adj, max_quartile)
# Defining x-y and stars coordinates
pairs_test_x_y <- pairs_test %>% 
  dplyr::select(taxon, group1, group2, p.value, max_quartile) %>% 
  mutate(pos = as.numeric(taxon),
         y = if_else(
           group1 == "Wildebeest", pos + 0.2, pos),
         yend = if_else(group2 == "Buffalo", pos, pos - 0.2),
         x = case_when(
           group1 == "Wildebeest" &
             group2 == "Buffalo" ~ max_quartile * 1.3),
           # group1 == "Wildebeest" &
           #   group2 == "Zebra" ~ max_quartile * 1.6,
           # group1 == "Wildebeest" &
           #   group2 == "Warthog" ~ max_quartile * 1.9),
         xend = x,
         x_star = 1.15*x,
         y_star = case_when(
           group1 == "Wildebeest" &
             group2 == "Buffalo" ~ pos + 0.05))
           # group1 == "Wildebeest" &
           #   group2 == "Zebra" ~ pos + 0.05,
           # group1 == "Wildebeest" &
           #   group2 == "Warthog" ~ pos + 0.05))
pairs_test %>%
  mutate(taxon = str_replace_all(taxon, "\\*", ""),
         taxon = str_replace_all(taxon, "<br>", " ")) %>% 
  saveRDS(., "RDataRDS/pairs_test.rds")
# Add significant stars to the otuplot
print(otuplot +
        geom_segment(data = pairs_test_x_y, aes(x = x, xend = xend, y = y,  yend = yend),
               inherit.aes = F) +
        geom_text(data = pairs_test_x_y,
            aes(x = x_star, y = y_star), label = "*", inherit.aes = F) +
        labs(subtitle = "Point range plot with \nstatistically significant stars"))
# Significant pairs
print(readRDS("RDataRDS/pairs_test.rds") %>% 
  dplyr::select(-max_quartile, 
         Taxon = taxon, 
         Group1 = group1, 
         Group2 = group2,
         P_value = p.value,
         P_adj = p_adj,
         ))
# Print error message if not significant.
} else {
cat("Sorry, no significant values found at 'P ≤ 0.05'\n", 
paste("P-values range from:",
      min(expt_signif$p.value), "-",
      max(expt_signif$p.value)))
}
```


## LEfSe: Linear discriminant analysis Effect Size {#lefse-analysis}
> LEfSe determines the significant differences between variables.

```{r lefse_fig}
source("R/lefse.R")
load("../iMAP-part3/RDataRDS/Rjoined_objects.RData", verbose = T)

metadata <- readRDS("../iMAP-part1/RDataRDS/mo_metadata.rds") %>% 
  dplyr::select(sample_id, isolate) %>% 
  drop_na(isolate)
shared_file <- read_tsv("~/Dropbox/CDILLC/GIT_REPOS/smda-end2end/data/final.tx.1.subsample.shared", show_col_types = F)
  
shared_design <- inner_join(shared_file, metadata, by=c("Group" = "sample_id"))
run_lefse("Buffalo", "Wildebeest", "buff_wilde")

read_tsv("RDataRDS/mothur_out/bush_buff_wilde.1.lefse_summary", show_col_types = F) %>% 
  mutate(OTU = tolower(OTU)) %>% 
  drop_na(LDA) %>% 
  inner_join(., Rjoined_sub_object, by = c("OTU" = "otu")) %>% 
  filter(level == "genus") %>% 
  mutate(LDA = if_else(Class == "Buffalo", -1 * LDA, LDA),
         taxon_phylo = fct_reorder(taxon_phylo, LDA)) %>% 
  ggplot(aes(x = LDA, y = taxon_phylo, fill = Class)) +
  geom_col() +
  labs(x = "LDA Score (log 10)", y = NULL) +
  scale_x_continuous(limits = c(-50, 25), breaks = seq(-50, 25, by=10)) +
  scale_fill_manual(name = NULL,
                  breaks = c("Buffalo",
                             "Wildebeest"),
                  labels = c("Variable 1",
                             "Variable 2"),
                  values = c("blue4",
                             "green4")) +
  theme_classic() +
  theme(
    axis.text.y = element_markdown(),
    legend.text = element_markdown()
  )
print(read_tsv("RDataRDS/mothur_out/bush_buff_wilde.1.lefse_summary", show_col_types = F) %>% 
  filter(pValue <= 0.05))
```

```{bash}
rm -f ./mothur*.logfile
```


> Lefse plot does not show much of the variation. We can create more plot like point range that shows the relative abundance, median etc.


# Alpha  Hypothesis Testing
- Assess the variation in alpha-diversity between groups.
- Then accept or reject the null hypothesis that there is no difference in location between groups.
- Box plots can show the location of the median.
- Perform samples Wilcoxon test.

## Descriptive statistics
```{r}
set.seed(1234)

readRDS("RDataRDS/ps_alpha_div_tbl.rds") %>%
  pivot_wider(id_cols = sample_id, names_from = metric, values_from = value) %>% 
  group_by(sample_id) %>%
    dplyr::summarise(
      median_observed = median(sobs),
      median_invsimpson = median(invsimpson),
      median_shannon = median(shannon)
      )
```

## Paired Wilcoxon test

### Observed vs Group variable
```{r}
bufwilde <- count_tbl %>% 
  filter(variable == "Buffalo"|
         variable == "Wildebeest")
         
  wilcox.test(bufwilde$sobs, exact = TRUE, conf.int = TRUE)

```

### Shannon vs Group
```{r}
wilcox.test(bufwilde$shannon, exact = FALSE, conf.int = TRUE)
```

### InvSimpson vs Group
```{r}
wilcox.test(bufwilde$invsimpson, exact = FALSE, conf.int = TRUE)
```

### Phylo Diversity vs Group
```{r}
wilcox.test(bufwilde$phylodiv, exact = FALSE, conf.int = TRUE)
```

## Breakaway richness estimates
- Estimate richness using breakaway
- Betta function tests for heterogeneity of total diversity (observed plus unobserved) across multiple sites. 
- It can account or test for fixed effects that may explain diversity. 
- It returns the significance of the covariates in explaining diversity and a hypothesis test for heterogeneity.
```{r}
library(breakaway)
ba_adiv <- breakaway::breakaway(ps)
ba_adiv[1]

#Plot estimates
plot(ba_adiv, ps, color = "isolate")     

#Examine models
summary(ba_adiv) %>%
  add_column("Group" = ps %>% otu_table %>% sample_names)  

# Test for group differ
bt <- breakaway::betta(summary(ba_adiv)$estimate, summary(ba_adiv)$error, make_design_matrix(ps, "isolate"))
bt$table
```


# Beta Hypothesis Testing

## Load and explore phyloseq objects
```{r}
load("../iMAP-part3/RDataRDS/phyloseq_objects.RData", verbose = T)
```

### Sample sums
```{r}
sample_sums(ps_raw) 

cat("\nSample sums range (lowest highest)\n")
sample_sums(ps_raw) %>% range
```

###  Perform multiple Rarefaction
- Sample 20 sequences from each sample.
- Repeat the procedure 10 times.

```{r}
samp100 <- phyloseq_mult_raref(ps_raw, SampSize = 100, iter = 10)
sample_sums(samp100[[1]])  # rarefied data
```

### Unifrac Sample dissimilarity
- Estimate sample dissimilarity independently for each iteration

```{r include=FALSE}
samp100_dis <- mult_dissim(samp100, method = "unifrac", average = F)
samp100_dis[[1]]
```

### Unifrac Average sample dissimilarities
```{r include=FALSE}
samp100_dis_avg <- mult_dissim(samp100, method = "unifrac", average = T)
samp100_dis_avg 
```

## ADONIS test

```{r}
vegan::adonis2(samp100_dis_avg ~ phyloseq::sample_data(ps_raw)$isolate)
```

## Homogeneity: Dispersion test
```{r}
dispr <- vegan::betadisper(samp100_dis_avg, phyloseq::sample_data(ps_raw)$isolate)
plot(dispr, main = "Homogeneity of multivariate dispersions", sub = "Ordination Centroids and Dispersion. ")
```

## Dispersion Boxplot
```{r}
boxplot(dispr, main = "Dispersion Boxplot", xlab = "Group")
```

## PERMANOVA: Permutation test for homogeneity
- Use the permutational multivariate analysis of variance (PERMANOVA)
- Accept or reject the null hypothesis.

